Secure Composition of Robust and Optimising Compilers

### 2023-06-22 popl/main.tex
Add sectioning

For languages used in this paper, a small-step style semantics is used.
That is, there is a step relation between states $\runtimetermvar$ and an event $\event$: $\stepto[]{\runtimetermvar[_{1}]}{\runtimetermvar[_{2}]}{\event}$.
Also, a program's syntax $p$ is encoded in the state $\runtimetermvar$ which is written as $p\in\runtimetermvar$.
For states $\runtimetermvar$, it is assumed that there exists a judgement $\isval{\runtimetermvar}$ that determines wether $\runtimetermvar$ is a final state or not.
\Cref{fig:starstep:and:nstep}\MK{appendix?} defines the star-step relation $\stepto{\runtimetermvar[_{1}]}{\runtimetermvar[_{2}]}{\trace}$, essentially modelling a big-step style evaluation, and the $n$-step relation $\stepton{\runtimetermvar[_{1}]}{\runtimetermvar[_{2}]}{\trace}$ that evaluates a program for $n$ steps.
Both relations filter $\varepsilon$ events.

\begin{figure}[t]
  \judgbox{\stepto{r}{r'}{\trace}}{,,State $\runtimetermvar$ steps to final state $\runtimetermvar[']$, emitting events $\trace$.''}$\;$\\
  %
  \typerule{star-step-refl}{
    \isval{\runtimetermvar}
  }{
    \stepto{\runtimetermvar}{\runtimetermvar}{\hole{\cdot}}
  }{star-step-refl}
  %
  \typerule{star-step-trans}{
    \event\not=\varepsilon &
    \stepto[]{\runtimetermvar}{\runtimetermvar[']}{\event} &
    \stepto{\runtimetermvar[']}{\runtimetermvar['']}{\trace}
  }{
    \stepto{\runtimetermvar}{\runtimetermvar['']}{\event\cdot\trace}
  }{star-step-trans}
  %
  \typerule{star-step-ign}{
    \stepto[]{\runtimetermvar}{\runtimetermvar[']}{\varepsilon} &
    \stepto{\runtimetermvar[']}{\runtimetermvar['']}{\trace}
  }{
    \stepto{\runtimetermvar}{\runtimetermvar['']}{\trace}
  }{star-step-ign}

  \judgbox{\stepton{\runtimetermvar}{\runtimetermvar[']}{\trace}}{,,State $r$ does $n$ steps to final state $\runtimetermvar[']$, emitting events $\trace$.''}$\;$\\
  %
  \typerule{n-step-refl}{
  }{
    \stepton[0]{\runtimetermvar}{\runtimetermvar}{\hole{\cdot}}
  }{n-step-refl}
  %
  \typerule{n-step-trans}{
    \event\not=\varepsilon &
    \stepto[]{\runtimetermvar}{\runtimetermvar[']}{\event} &
    \stepton{\runtimetermvar[']}{\runtimetermvar['']}{\trace}
  }{
    \stepton[n+1]{\runtimetermvar}{\runtimetermvar['']}{\event\cdot\trace}
  }{n-step-trans}
  %
  \typerule{n-step-ign}{
    \stepto[]{\runtimetermvar}{\runtimetermvar[']}{\varepsilon} &
    \stepton{\runtimetermvar[']}{\runtimetermvar['']}{\trace}
  }{
    \stepton[n+1]{\runtimetermvar}{\runtimetermvar['']}{\trace}
  }{n-step-ign}

  \caption{Star-step and $n$-step relation for a given step relation $\stepto{\runtimetermvar[_{1}]}{\runtimetermvar[_{2}]}{\event}$}\label{fig:starstep:and:nstep}
\end{figure}
%
The \emph{behavior} of a whole program $w$ ($\behav{w}$) is a set containing all traces the program may emit during execution.
Note that for deterministic languages, $\behav{w}$ is a singleton set for any $w$.
The behavior of programs gives rise to the definition of satisfaction:

@@

      \draw[->] (S) to[sloped,above] node {\text{\scriptsize\gls{tms}}} (T);
      \draw[->] (T) to[sloped,above] node {\text{\scriptsize\gls{sms}}} (M);
      \draw[->] (M) to[sloped,above] node {\text{\scriptsize\gls{dce}}} (D0);
      \draw[->] (M) to[sloped,above] node {\text{\scriptsize\gls{cf}}} (C0);
      \draw[->] (D0) to[sloped,above] node {\text{\scriptsize\gls{cf}}} (D1);
      \draw[->] (C0) to[sloped,above] node {\text{\scriptsize\gls{dce}}} (C1);
      \draw[->] (D1) to[sloped,above] node {\text{\scriptsize\gls{scct}}} (O);
      \draw[->] (C1) to[sloped,above] node {\text{\scriptsize\gls{scct}}} (O);
    \end{tikzpicture}
  \end{center}
  \caption{Visualisation of the optimizing compilation pipeline that attains a combination of \gls{ms} and \gls{cct}.}\label{fig:pipeline}
\end{figure}

@@

\subsection{$\src{L_{\tmssafe}}$: A Temporal but Not Spatial Memory Safe Language}\label{subsec:ltms}
\begin{gather*}
  \begin{aligned}
    \mi{(Pre\text{--}Types)}~\src{\type[_{e}]}&\bnfdef\src{\natt} \mid \src{\ptrn} \hspace{0.5cm}
    \mi{(Pointer\ Qualifiers)}~\src{\ptrqual}\bnfdef\ \src{\fullq} \mid \src{\halfq} \\
    \mi{(Type~Environment)}~\src{\typenv}&\bnfdef\ \src{\hole{\cdot}} \mid \hastype{\src{x}}{\src{\type}},\src{\typenv} \hspace{0.5cm}
    \mi{(Types)}~\src{\type}\bnfdef\src{\type[_{e}]} \mid \src{\type[_{e}]\to\bot} \mid \src{\type[^{(1)}_{e}]\to\type[^{(2)}_{e}]} \\
    \mi{(Symbols)}~\src{\function}&\bnfdef\src{\lfunction{g}{x:\type[_{e}]\to\type[_{e}]}{\expr}} \\
  \end{aligned}
\end{gather*}
\begin{center}
  \judgbox{\tcheck{\src{\typenv}}{\src{\expr}}{\src{\type}}}{,,Under environment $\src{\typenv}$, expression $\src{\expr}$ has type $\src{\type}$.''}
  %
  \typerule{t-$\src{x}$}{
    \src{\typenv}(\src{x})=\src{\type} &
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{x}}{\src{\type}}
  }{t-x}
  %
  \typerule{t-$\src{get}$}{
    \tcheck{\src{\typenv[_2]}}{\src{x}}{\src{\wptr}} &
    \tcheck{\src{\typenv[_1]}}{\src{\expr}}{\src{\natt}} &
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\lget{x}{\expr}}}{\src{\natt}}
  }{t-get}
  %
  \typerule{t-$\src{let}$}{
    \tcheck{\src{\typenv[_1]}}{\src{\expr[_1]}}{\src{\type[_e]}} &
    \tcheck{\hastype{\src{x}}{\src{\type[_e]}},\src{\typenv[_2]}}{\src{\expr[_2]}}{\src{\type}}
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\llet{x}{\expr[_1]}{\expr[_2]}}}{\src{\type}}
  }{t-let}
  %
  \typerule{t-$\src{delete}$}{
    \src{\typenv}(\src{x})=\src{\ptr} &
    \notowned{\src{\typenv}\setminus\left\{\hastype{\src{x}}{\src{\ptr}}\right\}}
  }{
    \tcheck{\src{\typenv}}{\src{\ldelete{x}}}{\src{\natt}}
  }{t-delete}
  %
  \typerule{t-$\src{abort}$}{
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{\labort}}{\src{\type}}
  }{t-abort}
  %
  \typerule{t-$\src{call}$}{
    \inttype{\src{\type[^{(1)}_e]}} &
    \inttype{\src{\type[^{(2)}_e]}} \\
    \tcheck{\src{\typenv}}{\src{foo}}{\src{\type[^{(1)}_e]\to\type[^{(2)}_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[^{(1)}_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lcall{foo}{\expr}}}{\src{\type[^{(2)}_e]}}
  }{t-call}
  %
  \typerule{t-$\src{return}$}{
    \inttype{\src{\type[_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lreturn{\expr}}}{\src{\type[_e]\to\bot}}
  }{t-return}
\end{center}

@@

\begin{gather*}
  \begin{aligned}
    \mi{(Expressions)}~\trg{\expr}\bnfdef\ \dots \mid \trg{\lispoisoned{x}} \mid \trg{\lpair{\expr[_{1}]}{\expr[_{2}]}} &\mid \trg{\lproja{\expr}} \mid \trg{\lprojb{\expr}} \mid \trg{\lhast{\expr}{\type}} \hspace{0.5cm}
    \mi{(Types)}~\trg{\type}\bnfdef\ \trg{\natt} \mid \trg{\natt\times\natt} \\
    \mi{(Values)}~\trg{\valueexpr}&\bnfdef\ \dots \mid \trg{\lpair{n_{1}}{n_{2}}}
  \end{aligned}
\end{gather*}
\begin{center}
  \typerule{e-$\trg{\lispoisoned{}}$-yes}{
    \trg{\statevar.\store}(\trg{x})=\trg{\tup{\loc}{\statevar.\sandboxtag,\poisoned,n}}
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lispoisoned{x}}}}{\rtt{\trg{\statevar}}{\trg{0}}}{\trg{\emptyevent}}
  }{e-ispoisoned-yes}
  %
  \typerule{e-$\trg{\lispoisoned{}}$-no}{
    \trg{\statevar.\store}(\trg{x})=\trg{\tup{\loc}{\statevar.\sandboxtag,\poisonless,n}}
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lispoisoned{x}}}}{\rtt{\trg{\statevar}}{\trg{1}}}{\trg{\emptyevent}}
  }{e-ispoisoned-no}
  %
  \typerule{e-proj1}{
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lproja{\lpair{n_1}{n_2}}}}}{\rtt{\trg{\statevar}}{\trg{n_1}}}{\trg{\emptyevent}}
  }{e-proj1}
  %
  \typerule{e-$\trg{\lhast{n}{\natt}}$}{
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lhast{n}{\natt}}}}{\rtt{\trg{\statevar}}{\trg{0}}}{\trg{\emptyevent}}
  }{e-n-has-nat}
  %
  \typerule{e-$\trg{\lhast{x}{\natt}}$}{
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lhast{x}{\natt}}}}{\rtt{\trg{\statevar}}{\trg{1}}}{\trg{\emptyevent}}
  }{e-x-has-nat}
  %
  \typerule{e-$\trg{\lhast{n}{\natt\times\natt}}$}{
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lhast{n}{\natt\times\natt}}}}{\rtt{\trg{\statevar}}{\trg{1}}}{\trg{\emptyevent}}
  }{e-n-has-nattimesnat}
\end{center}

The $\trg{L}$ language does not provide any guarantees with regards to memory safety.
The syntax of it extends the unified syntax for all languages (\Cref{subsec:cs:defs}) with pairs of natural numbers (\Cref{tr:e-proj1}) and two different kinds of dynamic checks.
One of those checks allows programmers to investigate wether a pointer is freed or not (\Cref{tr:e-ispoisoned-yes,tr:e-ispoisoned-no}).
The other check is a simple dynamic type check that evaluates to $\trg{0}$ if the check suceeds and to $\trg{1}$ otherwise (\Cref{tr:e-n-has-nat,tr:e-x-has-nat,tr:e-n-has-nattimesnat}).
As before, not all rules are shown due to their straightforwardness.
The projection $\filterL{\trace}$ is entirely similar to $\filterLtms{\trace}$.

@@

\begin{center}
  $$
  \begin{array}{rcl}
    \cca(\src{\lbinop{\expr[_{1}]}{\expr[_{2}]}}) & = & \lbinop[\trg]{\left[\cca(\src{\expr[_{1}]})\right]}{\left[\cca(\src{\expr[_{2}]})\right]} \\ [0.33cm]
    \cca(\src{\lfunction{g}{x:\natt\to\type[_{e}]}{\expr}}) & = & \lfunction[\trg]{\trg{g}}{\trg{x}}{\lifz[\trg]{\trg{\lhast{x}{\natt}}}{%
                                                                                                \left[\cca(\src{\expr})\right] %
                                                                                                 }{\labort[\trg]}}
  \end{array}
  $$
\end{center}

>>>>

For languages used in this paper, a small-step style semantics is used.
That is, there is a step relation between states $\runtimetermvar$ and an event $\event$: $\stepto[]{\runtimetermvar[_{1}]}{\runtimetermvar[_{2}]}{\event}$.
Also, a program's syntax $p$ is encoded in the state $\runtimetermvar$ which is written as $p\in\runtimetermvar$.
For states $\runtimetermvar$, it is assumed that there exists a judgement $\isval{\runtimetermvar}$ that determines wether $\runtimetermvar$ is a final state or not.
Another assumption is the existence of a generic star-step relation $\stepto{\runtimetermvar[_{1}]}{\runtimetermvar[_{2}]}{\trace}$, essentially modelling a big-step style evaluation, and the existence of an $n$-step relation $\stepton{\runtimetermvar[_{1}]}{\runtimetermvar[_{2}]}{\trace}$ that evaluates a program for $n$ steps.
Both relations filter $\varepsilon$ events.

The \emph{behavior} of a whole program $w$ ($\behav{w}$) is a set containing all traces the program may emit during execution.
Note that for deterministic languages, $\behav{w}$ is a singleton set for any $w$.
The behavior of programs gives rise to the definition of satisfaction:

@@

      \draw[->] (S) to[sloped,above] node (tmsedge) {\text{\scriptsize\gls{tms}}} (T);
      \draw[->] (T) to[sloped,above] node {\text{\scriptsize\gls{sms}}} (M);
      \draw[->] (M) to[sloped,above] node {\text{\scriptsize\gls{dce}}} (D0);
      \draw[->] (M) to[sloped,above] node {\text{\scriptsize\gls{cf}}} (C0);
      \draw[->] (D0) to[sloped,above] node {\text{\scriptsize\gls{cf}}} (D1);
      \draw[->] (C0) to[sloped,above] node {\text{\scriptsize\gls{dce}}} (C1);
      \draw[->] (D1) to[sloped,above] node {\text{\scriptsize\gls{scct}}} (O);
      \draw[->] (C1) to[sloped,above] node {\text{\scriptsize\gls{scct}}} (O);

      % Sections
      \node[above=1.5 of tmsedge] (sectms) {{\scriptsize\Cref{subsec:cs:tms}}};
      \node[right=0.0 of sectms] (secsms) {{\scriptsize\Cref{subsec:cs:ms}}};
      \node[right=1.25 of secsms] (secopts) {{\scriptsize\Cref{subsec:cs:opts}}};
      \node[right=1.25 of secopts] (secscct) {{\scriptsize\Cref{subsec:cs:scct}}};

      \draw[dashed] ($(sectms)-(0.7,0)$) -- ($(sectms)-(0.7,1.75)$);
      \draw[dashed] ($(sectms)-(-0.7,0)$) -- ($(sectms)-(-0.7,1.75)$);
      \draw[dashed] ($(secsms)-(-0.7,0)$) -- ($(secsms)-(-0.7,1.75)$);
      \draw[dashed] ($(secscct)-(0.7,0)$) -- ($(secscct)-(0.7,0.75)$);
      \draw[dashed] ($(secscct)-(-0.7,0)$) -- ($(secscct)-(-0.7,0.75)$);
    \end{tikzpicture}
  \end{center}
  \caption{Visualisation of the optimizing compilation pipeline that attains a combination of \gls{ms} and \gls{cct}.}\label{fig:pipeline}
\end{figure}

@@

\subsection{$\src{L_{\tmssafe}}$: A Temporal but Not Spatial Memory Safe Language}\label{subsec:ltms}
\begin{gather*}
  \begin{aligned}
    \mi{(Pre\text{--}Types)}~\src{\type[_{e}]}\bnfdef&\ \src{\natt} \mid \src{\ptrn} &
    \mi{(Type~Environment)}~\src{\typenv}\bnfdef&\ \src{\hole{\cdot}} \mid \hastype{\src{x}}{\src{\type}},\src{\typenv} \\
    \mi{(Types)}~\src{\type}\bnfdef&\ \src{\type[_{e}]} \mid \src{\type[_{e}]\to\bot} \mid \src{\type[^{(1)}_{e}]\to\type[^{(2)}_{e}]} &
    \mi{(Pointer\ Qualifiers)}~\src{\ptrqual}\bnfdef&\ \src{\fullq} \mid \src{\halfq} \\
    \mi{(Symbols)}~\src{\function}\bnfdef&\src{\lfunction{g}{x:\type[_{e}]\to\type[_{e}]}{\expr}} \\
  \end{aligned}
\end{gather*}
\begin{center}
  \judgbox{\tcheck{\src{\typenv}}{\src{\expr}}{\src{\type}}}{,,Under environment $\src{\typenv}$, expression $\src{\expr}$ has type $\src{\type}$.''}
  %
  \typerule{t-$\src{x}$}{
    \src{\typenv}(\src{x})=\src{\type} &
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{x}}{\src{\type}}
  }{t-x}
  %
  \typerule{t-$\src{get}$}{
    \tcheck{\src{\typenv[_2]}}{\src{x}}{\src{\wptr}} &
    \tcheck{\src{\typenv[_1]}}{\src{\expr}}{\src{\natt}} &
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\lget{x}{\expr}}}{\src{\natt}}
  }{t-get}
  %
  \typerule{t-$\src{let}$}{
    \tcheck{\src{\typenv[_1]}}{\src{\expr[_1]}}{\src{\type[_e]}} &
    \tcheck{\hastype{\src{x}}{\src{\type[_e]}},\src{\typenv[_2]}}{\src{\expr[_2]}}{\src{\type}}
  }{
    \tcheck{\typenvsplit{\src{\typenv[_1]}}{\src{\typenv[_2]}}}{\src{\llet{x}{\expr[_1]}{\expr[_2]}}}{\src{\type}}
  }{t-let}
  %
  \typerule{t-$\src{delete}$}{
    \src{\typenv}(\src{x})=\src{\ptr} &
    \notowned{\src{\typenv}\setminus\left\{\hastype{\src{x}}{\src{\ptr}}\right\}}
  }{
    \tcheck{\src{\typenv}}{\src{\ldelete{x}}}{\src{\natt}}
  }{t-delete}
  %
  \typerule{t-$\src{abort}$}{
    \notowned{\src{\typenv}}
  }{
    \tcheck{\src{\typenv}}{\src{\labort}}{\src{\type}}
  }{t-abort}
  %
  \typerule{t-$\src{call}$}{
    \inttype{\src{\type[^{(1)}_e]}} &
    \inttype{\src{\type[^{(2)}_e]}} \\
    \tcheck{\src{\typenv}}{\src{foo}}{\src{\type[^{(1)}_e]\to\type[^{(2)}_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[^{(1)}_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lcall{foo}{\expr}}}{\src{\type[^{(2)}_e]}}
  }{t-call}
  %
  \typerule{t-$\src{return}$}{
    \inttype{\src{\type[_e]}} &
    \tcheck{\src{\typenv}}{\src{\expr}}{\src{\type[_e]}}
  }{
    \tcheck{\src{\typenv}}{\src{\lreturn{\expr}}}{\src{\type[_e]\to\bot}}
  }{t-return}
\end{center}

@@

\begin{gather*}
  \begin{aligned}
    \mi{(Expressions)}~\trg{\expr}\bnfdef\ \dots \mid \trg{\lispoisoned{x}} \mid \trg{\lpair{\expr[_{1}]}{\expr[_{2}]}} &\mid \trg{\lproja{\expr}} \mid \trg{\lprojb{\expr}} \mid \trg{\lhast{\expr}{\type}} \hspace{0.5cm}
    \mi{(Types)}~\trg{\type}\bnfdef\ \trg{\natt} \mid \trg{\natt\times\natt} \\
    \mi{(Values)}~\trg{\valueexpr}&\bnfdef\ \dots \mid \trg{\lpair{n_{1}}{n_{2}}}
  \end{aligned}
\end{gather*}
\begin{center}
  \typerule{e-$\trg{\lispoisoned{}}$-yes}{
    \trg{\statevar.\store}(\trg{x})=\trg{\tup{\loc}{\statevar.\sandboxtag,\poisoned,n}}
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lispoisoned{x}}}}{\rtt{\trg{\statevar}}{\trg{0}}}{\trg{\emptyevent}}
  }{e-ispoisoned-yes}
  %
%  \typerule{e-$\trg{\lispoisoned{}}$-no}{
%    \trg{\statevar.\store}(\trg{x})=\trg{\tup{\loc}{\statevar.\sandboxtag,\poisonless,n}}
%  }{
%    \pstepto{\rtt{\trg{\statevar}}{\trg{\lispoisoned{x}}}}{\rtt{\trg{\statevar}}{\trg{1}}}{\trg{\emptyevent}}
%  }{e-ispoisoned-no}
  %
  \typerule{e-$\trg{\lhast{n}{\natt}}$}{
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lhast{n}{\natt}}}}{\rtt{\trg{\statevar}}{\trg{0}}}{\trg{\emptyevent}}
  }{e-n-has-nat}
  %
%  \typerule{e-$\trg{\lhast{x}{\natt}}$}{
%  }{
%    \pstepto{\rtt{\trg{\statevar}}{\trg{\lhast{x}{\natt}}}}{\rtt{\trg{\statevar}}{\trg{1}}}{\trg{\emptyevent}}
%  }{e-x-has-nat}
  %
  \typerule{e-$\trg{\lhast{n}{\natt\times\natt}}$}{
  }{
    \pstepto{\rtt{\trg{\statevar}}{\trg{\lhast{n}{\natt\times\natt}}}}{\rtt{\trg{\statevar}}{\trg{1}}}{\trg{\emptyevent}}
  }{e-n-has-nattimesnat}
\end{center}

The $\trg{L}$ language does not provide any guarantees with regards to memory safety.
The syntax of it extends the unified syntax for all languages (\Cref{subsec:cs:defs}) with pairs of natural numbers (\Cref{tr:e-proj1}) and two different kinds of dynamic checks.
One of those checks allows programmers to investigate wether a pointer is freed or not (\Cref{tr:e-ispoisoned-yes}).
The other check is a simple dynamic type check that evaluates to $\trg{0}$ if the check suceeds and to $\trg{1}$ otherwise (\Cref{tr:e-n-has-nat,tr:e-n-has-nattimesnat}).
The projection $\filterL{\trace}$ is entirely similar to $\filterLtms{\trace}$.

@@

\begin{center}
  $$
  \begin{array}{rcl}
%    \cca(\src{\lbinop{\expr[_{1}]}{\expr[_{2}]}}) & = & \lbinop[\trg]{\left[\cca(\src{\expr[_{1}]})\right]}{\left[\cca(\src{\expr[_{2}]})\right]} \\ [0.33cm]
    \cca(\src{\lfunction{g}{x:\natt\to\type[_{e}]}{\expr}}) & = & \lfunction[\trg]{\trg{g}}{\trg{x}}{\lifz[\trg]{\trg{\lhast{x}{\natt}}}{%
                                                                                                \left[\cca(\src{\expr})\right] %
                                                                                                 }{\labort[\trg]}}
  \end{array}
  $$
\end{center}
